---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

Unzip the input zip file to the current directory, read the csv file and convert dates to the R date format:
```{r preprocessing}
unzip("activity.zip",exdir="./")
activity<-read.csv("activity.csv",stringsAsFactor=FALSE)
activity$date<-as.Date(activity$date,format="%Y-%m-%d")
```



## What is mean total number of steps taken per day?

Group data frame by dates and calculate sum of steps taken on given day:
```{r group_by_date}
library(dplyr) # to help manipulate dataset
activity_by_date<-group_by(activity,date)
total_steps<-summarise(activity_by_date,sum(steps))
```

Make a numeric vector of total number of steps taken every day:
```{r make_step_vector}
steps_per_day<-total_steps[[2]]
```

Create a histogram with distribution of the total number of steps per day and save it to the file:
```{r make_histogram,fig.height=4,fig.path="figure/",}
hist(steps_per_day,xlab="Total number of steps per day",main="",col = "red")
```

Now calculate the mean and median:
```{r mean_total_steps}
mean(steps_per_day[!is.na(steps_per_day)])
median(steps_per_day[!is.na(steps_per_day)])
```


## What is the average daily activity pattern?

Average the intervals over all days (and exclude the missing values from averaging):
```{r group_by_interval}
activity_by_interval<-group_by(activity,interval) #group by interval
avg_steps_interval<-summarise(activity_by_interval,mean(steps[!is.na(steps)]))
```

Make a time series plot of average steps taken in a given interval during day. Note that the times are shown without separation between hours and minutes and are in 24h format (e.g. 500 is 5am and 1500 is 3pm).

```{r make_time_avg,fig.height=4,fig.path="figure/",}
 plot(avg_steps_interval,type="l",ylab="average number of steps",
      main="Average number of steps in an interval over the observed period",
      xlab="Time during the day")
```

Finding which interval contains on average the highest number of steps:
```{r find_max}
row_id<-which.max(avg_steps_interval[[2]]) # find on which line there is maximum
avg_steps_interval$interval[row_id] # get the interval
```
So the maximum number of steps was on average taken around **8:35** in the morning

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
